\bf A note on the instability and degeneracy of deep learning models

### 2017-11-06 writing/paper-note/docs/note.tex
ready?

\title{\bf A Note on the Instability and Degeneracy of Deep Learning Models}
\author{Andee Kaplan \\ Department of Statistics, Iowa State University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
Manufacturing Systems Engineering, Iowa State University \\ }

\maketitle
\begin{abstract}
A probability model exhibits instability if small changes in a data
outcome result in large, and often unanticipated, changes in
probability. For correlated data structures found in several application
areas, there is increasing interest in predicting/identifying such
sensitivity in model probability structure. We consider the problem of
quantifying instability for general probability models defined on
sequences of observations, where each sequence of length \(N\) has a
finite number of possible outcomes. A sequence of probability models
results, indexed by \(N\), that accommodates data of expanding
dimension. Model instability is formally shown to occur when a certain
log-probability ratio under such models grows faster than \(N\). In this
case, a one component change in the data sequence can shift probability
by orders of magnitude. Also, as instability becomes more extreme, the
resulting probability models are shown to tend to degeneracy, placing
all their probability on potentially small portions of the sample space.
These results on instability apply to large classes of models commonly
used in random graphs, network analysis, and machine learning contexts.
\end{abstract}
\noindent%
{\it Keywords:}  Degeneracy, Instability, Classification, Deep Learning, Graphical Models
\vfill

@@

The above model condition incorporates many standard parameterizations
and follows, for instance, whenever
\(P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y) = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}\)
holds for outcomes \(\boldsymbol x, \boldsymbol y \in\mathcal{X}^N\) in
a FOES model. For instance, this latter condition is fulfilled for all
linear exponential families from Section
\ref{discrete-exponential-family-models} (e.g.,
\eqref{eq:mod1}-\eqref{eq:mod2}) as well as all network models from Sections
\ref{restricted-boltzmann-machines}-\ref{deep-learning} (e.g.,
\eqref{eq:RBM1}-\eqref{eq:RBM2}). When parameters can be tuned in sign with
effects prescribed in the model condition PSR, unstable FOES models will
exhibit further probability sensitivities, as outlined in the following
extension of Theorem @ref(thm:degenFOES\}).

\BeginKnitrBlock{corollary}
\protect\hypertarget{cor:sign}{}{\label{cor:sign}}Let
\(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\),
\(N\geq 1\), be a sequence of FOES models satisfying model condition
PSR. If the models \(P_{\boldsymbol \theta_N}\) are additionally
S-unstable, then
\begin{enumerate}[(i)]
\item the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
\item and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $- \boldsymbol \theta_N$ that
    $$ 
    P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \mbox{as $N\to \infty$},
    $$
    while, by Theorem \ref{thm:degenFOES}, $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for $\boldsymbol X_N = (X_1,\ldots,X_N)$ under $\boldsymbol \theta_N$.
\end{enumerate}
\EndKnitrBlock{corollary}

@@

\BeginKnitrBlock{remark}
\iffalse <span class="remark"><em>Remark. \fi{} Model condition PSR and
Corollary \ref{cor:sign} can be extended to cases where parameter
components
\(\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})\)
are not all changed in sign (\(-\boldsymbol \theta_N\)) but, more
generally, are instead altered to another parameter configuration
\(\boldsymbol \theta_{N}^A = (\boldsymbol \theta_{N,1}^A, \boldsymbol \theta_{N,2}^A)\)
involving a switch in sign only among some model parameters
\(\boldsymbol \theta_{N,2}^A = -\boldsymbol \theta_{N,2}\) with
remaining parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily
chosen. If a sign change occurs among parameters
(\(\pm \boldsymbol \theta_{N,2}\)) which dominate the probability
structure of the model, then an unstable model under
\(\boldsymbol \theta_N\) will then imply that many more unstable models
exist over a broad spectrum of possibilities for variations
\(\boldsymbol \theta_{N}^A\) of \(\boldsymbol \theta_N\).
\EndKnitrBlock{remark}

@@

A fundamental model considered in the instability work of Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) involves a
one-parameter exponential model corresponding to \eqref{eq:expo} with a
real-valued parameter, say
\(\boldsymbol \theta_N = \eta(\boldsymbol \theta_N)\in \mathbb{R}\), and
sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}\).
For such models, upon scaling by sample size \(N\), the log-ratio of
extreme probabilities in \eqref{eq:elpr} for assessing instability becomes
\begin{equation}
\label{eq:UL}
\frac{1}{N}\REP(\boldsymbol \theta_N ) \equiv   |\boldsymbol \theta_N| \frac{(U_N-L_N)}{N} \;\quad \mbox{(one-parameter exponential model)},
\end{equation}
where
\(U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
and
\(L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
denote the maximal and minimal values of the single sufficient
statistic. In this case, an S-unstable model results, by definition
\eqref{eq:Sun}, whenever
\(\lim_{N\to \infty} |\boldsymbol \theta_N| (U_N-L_N)/N= \infty\) holds
or, in other words, if the combined magnitudes of parameter
\(|\boldsymbol \theta_N|\) and maximal difference \(U_N-L_N\) in
statistic values are overwhelmingly large relative to the sample size
\(N\). If we further assume that
\(\boldsymbol \theta_N =\theta\in\mathbb{R}\setminus \{0\}\) is a fixed
(non-zero) parameter for all \(N \geq 1\), as considered in
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), then an
S-unstable model results solely if the sufficient statistic admits a
value \(U_N-L_N\) too large relative to number \(N\) of observations,
i.e., if \((U_N-L_N)/N\to \infty\) as \(N\to \infty\). The latter aspect
reflects the definition of Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), for this
setting, that a real-valued \emph{statistic} \(g_N(\boldsymbol x)\) may
be classified as \emph{unstable} when
\(\lim_{N\to \infty}|(U_N-L_N)/N=\infty\) holds and as \emph{stable}
otherwise (e.g., if \(\sup_{N \geq 1}(U_N-L_N)/N<\infty\)).

For illustration, consider the iid Bernoulli model \eqref{eq:mod1} for
\(\boldsymbol X_N=(X_1,\ldots,X_N)\) with log-odds ratio parameter
\(\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}\).
Remark 1 (Section \ref{criterion}) then gives the model instability
measure \eqref{eq:Sun} directly as \[
\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \mbox{(iid Bernoulli model)},
\] so that an unstable (or stable) model results for a divergent (or
bounded) parameter sequence \(|\boldsymbol \theta_N|\). The above
instability expression for the Bernoulli model follows as well from the
\(N\)-scaled LREP value \eqref{eq:UL} for a one-parameter exponential
distribution, using that the sufficient statistic involved
\(g_N(\boldsymbol x)= \sum_{i=1}^N x_i\),
\(\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N\), has maximum and mimimum
values \(U_N=N\) and \(L_N=0\). In this case, Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) has noted
that the statistic is stable (i.e., bounded \((U_N-L_N)/N=1\)) and the
Bernoulli model is as well when, in particular,
\(\boldsymbol \theta_N=\bold \theta \in\mathbb{R}\) is fixed for
\(N \geq 1\).

Alternatively, considering a random graph with \(N={n \choose 2}\) edges
among \(n\) nodes, the exponential graph model from \eqref{eq:mod2}, when
based purely on the number of \(g_{2,N}(\boldsymbol x)\) of 2-stars or
solely the number \(g_{3,N}(\boldsymbol x)\) of triangles,
\(\boldsymbol x\in\{0,1\}^N\), has an measure of instability from
\eqref{eq:Sun} as \[
\frac{1}{N}\REP(\boldsymbol \theta_N )  = \left\{ \begin{array}{lcl} |\boldsymbol \theta_N| (n-2) && \mbox{(2-star graph model)}\\
|\boldsymbol \theta_N|(n-2)/3 &&\mbox{(triangle graph model)},\end{array}\right.
\] by using the (one-parameter exponential) LREP formula \eqref{eq:UL}
with statistic maximums \(U_N= N(n-2)\) for 2-stars or \(U_N= N(n-2)/3\)
for triangles and with minimums \(L_N=0\) in both cases. Because the
variable number \(N\to \infty\) as the node number \(n\to \infty\), both
counts of 2-stars and triangles are unstable statistics in the sense of
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) (i.e.,
\(\lim_{N\to \infty} (U_N-L_N)/N=\infty\)). Furthermore, both types of
graph models are always S-unstable for all possible of parameter
sequences \(\boldsymbol \theta_N \in\mathbb{R}\) that are bounded away
from zero (i.e.,
\(\lim_{N\to \infty}\REP(\boldsymbol \theta_N )/N=\infty\) then holds,
including the fixed parameter case
\(\boldsymbol \theta_N=\theta\in\mathbb{R}\setminus \{0\}\) from
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011})).

@@

Proposition \ref{prp:prop1} provides a sufficient condition for the
stability of linear exponential models with fixed parameter dimension
\(k\geq 1\), whereby an S-stable model is guaranteed if the compounded
magnitude of each combination of parameter \(\theta_{i,N}\) and
sufficient statistic value \((U_{i,N}-L_{i,N})\) is bounded by the
sample size \(N\), \(i=1,\ldots,k\). This supports the findings of
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), who showed
degeneracy follows in such models under one type of violation of the
condition \eqref{eq:prop1} in Proposition \ref{prp:prop1} (namely,
involving \(k>1\) non-zero parameters with \(k-1\) statistics being
\(O(N)\) bounded while one statistic diverges in maximal size faster
than the number \(N\) of observations). To further illustrate the result
in Proposition \ref{prp:prop1}, consider the multinomial distribution
\eqref{eq:mod11} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) having
\(k\geq 2\) categories \(\{1,\ldots,k\}\) and \(k\) parameters
\(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\).
The variables are iid under this model so that Remark 1 (Section
\ref{criterion} yields the corresponding \(N\)-scaled log-ratio of
extreme probabilities \eqref{eq:elpr} as
\begin{align*}
  \frac{1}{N}\REP(\boldsymbol \theta_N) &= \frac{\max_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}{\min_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}\\
  &= \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \mbox{(iid multinomial model)}.
\end{align*}
Hence, a multinomial model sequence is unstable (or stable) depending on
whether (or not) the maximal parameter difference
\(\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\)
diverges. Furthermore, using that each of the \(k\) sufficient (count)
statistics from the multinomial model \eqref{eq:mod11} satisfies
\((U_{i,N}-L_{i,N})/N=1\), we see that \eqref{eq:prop1} of Proposition
\ref{prp:prop1} becomes purely a parameter condition,
\(\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,N}| <\infty\), for
ensuring that
\(\REP(\boldsymbol \theta_N)/N =\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\)
is bounded and stability follows for the multinomial distribution.
Additionally, a stable multinomial sequence (i.e., bounded
\(\REP(\boldsymbol \theta_N)/N\)) turns out to be nearly equivalent to
\eqref{eq:prop1} (e.g., these are the same if the smallest parameter
\(\min_{1 \leq i \leq k } |\theta_{i,N}|\) remains bounded).

@@

We next consider instability of discrete data models based on
exponential formulations involving hidden, or latent, variables, such as
those probabilistic graphical models described in Sections
\ref{restricted-boltzmann-machines}-\ref{deep-learning}. We will focus
on restricted Bolzmann machine (RBM) models (Section
\ref{restricted-boltzmann-machines}, having one layer of latent
variables for simplicity, though the same instability concepts may be
extended to other deep learning models (Section \ref{deep-learning}. For
\(N\) visible variables
\(\boldsymbol X \equiv \boldsymbol X_N = (X_1,\ldots,X_n)\) as data,
each observation \(X_i\in\{\pm 1\}\) being binary, the RBM-based model
\eqref{eq:RBM2} for \(\boldsymbol X\) is again of FOES-type, though not an
exponential model. However, the distribution of visible variables is
induced by an underlying joint exponential model \eqref{eq:RBM1} for both
visible and latent variables \((\boldsymbol X, \boldsymbol H)\), where
\(\boldsymbol H = (H_1,\ldots,H_{N_{\mathcal{H}}})\) denotes a vector of
\(N_{\mathcal{H}}\) hidden variables (similarly binary). The joint model
is of linear exponential form involving
\(q(N)\equiv N + N_{\mathcal{H}} + N*N_{\mathcal{H}}\) sufficient
statistics given by
\((\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)\) and
parameters
\(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}} ) \in\mathbb{R}^{q(N)}\)
corresponding to the \(N\) visible variables \(\boldsymbol X\) (i.e.,
\(\boldsymbol \theta_N^{\mathcal{V}}\in\mathbb{R}^N\)), the
\(N_{\mathcal{H}}\) hidden variables \(\boldsymbol H\) (i.e.,
\(\boldsymbol \theta_N^{\mathcal{H}}\in\mathbb{R}^{N_{\mathcal{H}}}\)),
and the \(N *N_{\mathcal{H}}\) cross-product variables
\(\boldsymbol X^T\boldsymbol H\) (i.e.,
\(\boldsymbol \theta_N^{\mathcal{VH}}\in\mathbb{R}^{N*N_{\mathcal{H}}}\)).
However, unlike some previous exponential models considered in Sections
\ref{one-param-exp}-\ref{fixed-dim-exp} (cf.~Proposition
\ref{prp:prop1}, note that the RBM formulation always associates
parameters with \emph{bounded} statistics (i.e., the components of
\((\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)\)) so
that model instability cannot arise here due to the magnitude of
sufficient statistics exceeding the sample size \(N\). Instead, RBM
instability may be linked solely to parameter configuration and the fact
that the number \(q(N) \geq N\) of parameters necessarily increases with
the number \(N\) of observations \(\boldsymbol X\), in contrast to
previous exponential cases of fixed parameter dimension.

>>>>

\title{\bf A note on the instability and degeneracy of deep learning models}
\author{Andee Kaplan \\ Department of Statistics, Iowa State University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
Manufacturing Systems Engineering, Iowa State University \\ }

\maketitle
\begin{abstract}
A probability model exhibits instability if small changes in a data
outcome result in large, and often unanticipated, changes in
probability. This instability is a property of the probability model,
rather than the fitted parameter vector. For correlated data structures
found in several application areas, there is increasing interest in
predicting/identifying such sensitivity in model probability structure.
We consider the problem of quantifying instability for general
probability models defined on sequences of observations, where each
sequence of length \(N\) has a finite number of possible values. A
sequence of probability models results, indexed by \(N\), that
accommodates data of expanding dimension. Model instability is formally
shown to occur when a certain log-probability ratio under such models
grows faster than \(N\). In this case, a one component change in the
data sequence can shift probability by orders of magnitude. Also, as
instability becomes more extreme, the resulting probability models are
shown to tend to degeneracy, placing all their probability on
potentially small portions of the sample space. These results on
instability apply to large classes of models commonly used in random
graphs, network analysis, and machine learning contexts.
\end{abstract}
\noindent%
{\it Keywords:}  Degeneracy, Instability, Classification, Deep Learning, Graphical Models
\vfill

@@

The above model condition incorporates many standard parameterizations
and follows, for instance, whenever
\(P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y) = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}\)
holds for outcomes \(\boldsymbol x, \boldsymbol y \in\mathcal{X}^N\) in
a FOES model. For instance, this latter condition is fulfilled for all
linear exponential families from Section
\ref{discrete-exponential-family-models} (e.g.,
\eqref{eq:mod1}-\eqref{eq:mod2}) as well as all network models from Sections
\ref{rbm}-\ref{deep-learning} (e.g., \eqref{eq:RBM1}-\eqref{eq:RBM2}). When
parameters can be tuned in sign with effects prescribed in the model
condition PSR, unstable FOES models will exhibit further probability
sensitivities, as outlined in the following extension of Theorem
\ref{thm:degenFOES}.

\BeginKnitrBlock{corollary}
\protect\hypertarget{cor:sign}{}{\label{cor:sign} }Let
\(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\),
\(N\geq 1\), be a sequence of FOES models satisfying model condition
PSR. If the models \(P_{\boldsymbol \theta_N}\) are additionally
S-unstable, then
\begin{enumerate}[(i)]
\item the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
\item and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $-\boldsymbol \theta_N$ that
    $$ 
    P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \text{as } N\to \infty,
    $$
    while, by Theorem \ref{thm:degenFOES}, $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for $\boldsymbol X_N = (X_1,\ldots,X_N)$ under $\boldsymbol \theta_N$.
\end{enumerate}
\EndKnitrBlock{corollary}

@@

\BeginKnitrBlock{remark}
\iffalse{} {Remark. } \fi{}Under the model condition PSR, Corollary
\ref{cor:sign} can also be extended to cases where parameter components
\(\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})\)
(say) are not all changed in sign (e.g., \(-\boldsymbol \theta_N\)) but,
more generally, are instead altered to another parameter configuration
\(\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )\)
involving a switch in sign only among some dominating model parameters
\(\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}\) with remaining
parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily chosen. If a
change sign occurs among parameters (\(\pm \boldsymbol \theta_{N,2}\))
which dominate the probability structure of the model, then the results
of Corollary \ref{cor:sign} can still hold with
\(\boldsymbol \theta_{N}^A\) replacing \(-\boldsymbol \theta_N\); as an
example of one sufficient condition, if
\(\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0\)
holds in addition to Corollary \ref{cor:sign} assumptions, where \[
G_N(\boldsymbol x, \boldsymbol \theta) =   \frac{\log P_{\boldsymbol \theta}(\boldsymbol x) - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }{\max_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y)  - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }, \quad \boldsymbol x \in\mathcal{X}^N,
\] represents a standardized form of \(\boldsymbol \theta\)-model
probabilities, then the results of Corollary \ref{cor:sign} apply to
\(\boldsymbol \theta_{N}^A\) in addition to \(-\boldsymbol \theta_N\).
As a consequence, an unstable model under \(\boldsymbol \theta_N\) can
then imply that many more unstable models exist over a broader spectrum
of possibilities for variations \(\boldsymbol \theta_{N}^A\) of
\(\boldsymbol \theta_N\), which involves some amount of sign change
among components of \(\boldsymbol \theta_N\).
\EndKnitrBlock{remark}

@@

A fundamental model considered in the instability work of Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) involves a
one-parameter exponential model corresponding to \eqref{eq:expo} with a
real-valued parameter, say
\(\boldsymbol \theta_N = \eta(\boldsymbol \theta_N)\in \mathbb{R}\), and
sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}\).
For such models, upon scaling by sample size \(N\), the log-ratio of
extreme probabilities in \eqref{eq:elpr} for assessing instability becomes
\begin{equation}
\label{eq:UL}
\frac{1}{N}\REP(\boldsymbol \theta_N ) \equiv   |\boldsymbol \theta_N| \frac{(U_N-L_N)}{N} \;\quad \text{(one-parameter exponential model)},
\end{equation}
where
\(U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
and
\(L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
denote the maximal and minimal values of the single sufficient
statistic. In this case, an S-unstable model results, by definition
\eqref{eq:Sun}, whenever
\(\lim_{N\to \infty} |\boldsymbol \theta_N| (U_N-L_N)/N= \infty\) holds
or, in other words, if the combined magnitudes of parameter
\(|\boldsymbol \theta_N|\) and maximal difference \(U_N-L_N\) in
statistic values are overwhelmingly large relative to the sample size
\(N\). If we further assume that
\(\boldsymbol \theta_N =\theta\in\mathbb{R}\setminus \{0\}\) is a fixed
(non-zero) parameter for all \(N \geq 1\), as considered in
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), then an
S-unstable model results solely if the sufficient statistic admits a
value \(U_N-L_N\) too large relative to number \(N\) of observations,
i.e., if \((U_N-L_N)/N\to \infty\) as \(N\to \infty\). The latter aspect
reflects the definition of Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), for this
setting, that a real-valued \emph{statistic} \(g_N(\boldsymbol x)\) may
be classified as \emph{unstable} when
\(\lim_{N\to \infty}|(U_N-L_N)/N=\infty\) holds and as \emph{stable}
otherwise (e.g., if \(\sup_{N \geq 1}(U_N-L_N)/N<\infty\)).

For illustration, consider the iid Bernoulli model \eqref{eq:mod1} for
\(\boldsymbol X_N=(X_1,\ldots,X_N)\) with log-odds ratio parameter
\(\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}\).
Remark 1 (Section \ref{criterion}) then gives the model instability
measure \eqref{eq:Sun} directly as \[
\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \text{(iid Bernoulli model)},
\] so that an unstable (or stable) model results for a divergent (or
bounded) parameter sequence \(|\boldsymbol \theta_N|\). The above
instability expression for the Bernoulli model follows as well from the
\(N\)-scaled LREP value \eqref{eq:UL} for a one-parameter exponential
distribution, using that the sufficient statistic involved
\(g_N(\boldsymbol x)= \sum_{i=1}^N x_i\),
\(\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N\), has maximum and minimum
values \(U_N=N\) and \(L_N=0\). In this case, Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) has noted
that the statistic is stable (i.e., bounded \((U_N-L_N)/N=1\)) and the
Bernoulli model is as well when, in particular,
\(\boldsymbol \theta_N=\bold \theta \in\mathbb{R}\) is fixed for
\(N \geq 1\).

Alternatively, considering a random graph with \(N={n \choose 2}\) edges
among \(n\) nodes, the exponential graph model from \eqref{eq:mod2}, when
based purely on the number of \(g_{2,N}(\boldsymbol x)\) of 2-stars or
solely the number \(g_{3,N}(\boldsymbol x)\) of triangles,
\(\boldsymbol x\in\{0,1\}^N\), has an measure of instability from
\eqref{eq:Sun} as \[
\frac{1}{N}\REP(\boldsymbol \theta_N )  = \left\{ \begin{array}{lcl} |\boldsymbol \theta_N| (n-2) && \text{(2-star graph model)}\\
|\boldsymbol \theta_N|(n-2)/3 &&\text{(triangle graph model)},\end{array}\right.
\] by using the (one-parameter exponential) LREP formula \eqref{eq:UL}
with statistic maximums \(U_N= N(n-2)\) for 2-stars or \(U_N= N(n-2)/3\)
for triangles and with minimums \(L_N=0\) in both cases. Because the
variable number \(N\to \infty\) as the node number \(n\to \infty\), both
counts of 2-stars and triangles are unstable statistics in the sense of
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) (i.e.,
\(\lim_{N\to \infty} (U_N-L_N)/N=\infty\)). Furthermore, both types of
graph models are always S-unstable for all possible of parameter
sequences \(\boldsymbol \theta_N \in\mathbb{R}\) that are bounded away
from zero (i.e.,
\(\lim_{N\to \infty}\REP(\boldsymbol \theta_N )/N=\infty\) then holds,
including the fixed parameter case
\(\boldsymbol \theta_N=\theta\in\mathbb{R}\setminus \{0\}\) from
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011})).

@@

Proposition \ref{prp:prop1} provides a sufficient condition for the
stability of linear exponential models with fixed parameter dimension
\(k\geq 1\), whereby an S-stable model is guaranteed if the compounded
magnitude of each combination of parameter \(\theta_{i,N}\) and
sufficient statistic value \((U_{i,N}-L_{i,N})\) is bounded by the
sample size \(N\), \(i=1,\ldots,k\). This supports the findings of
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), who showed
degeneracy follows in such models under one type of violation of the
condition \eqref{eq:prop1} in Proposition \ref{prp:prop1} (namely,
involving \(k>1\) non-zero parameters with \(k-1\) statistics being
\(O(N)\) bounded while one statistic diverges in maximal size faster
than the number \(N\) of observations). To further illustrate the result
in Proposition \ref{prp:prop1}, consider the multinomial distribution
\eqref{eq:mod11} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) having
\(k\geq 2\) categories \(\{1,\ldots,k\}\) and \(k\) parameters
\(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\).
The variables are iid under this model so that Remark 1 (Section
\ref{criterion} yields the corresponding \(N\)-scaled log-ratio of
extreme probabilities \eqref{eq:elpr} as
\begin{align*}
  \frac{1}{N}\REP(\boldsymbol \theta_N) &= \frac{\max_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}{\min_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}\\
  &= \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \text{(iid multinomial model)}.
\end{align*}
Hence, a multinomial model sequence is unstable (or stable) depending on
whether (or not) the maximal parameter difference
\(\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\)
diverges. Furthermore, using that each of the \(k\) sufficient (count)
statistics from the multinomial model \eqref{eq:mod11} satisfies
\((U_{i,N}-L_{i,N})/N=1\), we see that \eqref{eq:prop1} of Proposition
\ref{prp:prop1} becomes purely a parameter condition,
\(\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,N}| <\infty\), for
ensuring that
\(\REP(\boldsymbol \theta_N)/N =\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\)
is bounded and stability follows for the multinomial distribution.
Additionally, a stable multinomial sequence (i.e., bounded
\(\REP(\boldsymbol \theta_N)/N\)) turns out to be nearly equivalent to
\eqref{eq:prop1} (e.g., these are the same if the smallest parameter
\(\min_{1 \leq i \leq k } |\theta_{i,N}|\) remains bounded).

@@

We next consider instability of discrete data models based on
exponential formulations involving hidden, or latent, variables, such as
those probabilistic graphical models described in Sections
\ref{rbm}-\ref{deep-learning}. We will focus on restricted Boltzmann
machine (RBM) models (Section \ref{rbm}, having one layer of latent
variables for simplicity, though the same instability concepts may be
extended to other deep learning models (Section \ref{deep-learning}. For
\(N\) visible variables
\(\boldsymbol X \equiv \boldsymbol X_N = (X_1,\ldots,X_n)\) as data,
each observation \(X_i\in\{\pm 1\}\) being binary, the RBM-based model
\eqref{eq:RBM2} for \(\boldsymbol X\) is again of FOES-type, though not an
exponential model. However, the distribution of visible variables is
induced by an underlying joint exponential model \eqref{eq:RBM1} for both
visible and latent variables \((\boldsymbol X, \boldsymbol H)\), where
\(\boldsymbol H = (H_1,\ldots,H_{N_{\mathcal{H}}})\) denotes a vector of
\(N_{\mathcal{H}}\) hidden variables (similarly binary). The joint model
is of linear exponential form involving
\(q(N)\equiv N + N_{\mathcal{H}} + N*N_{\mathcal{H}}\) sufficient
statistics given by
\((\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)\) and
parameters
\(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}} ) \in\mathbb{R}^{q(N)}\)
corresponding to the \(N\) visible variables \(\boldsymbol X\) (i.e.,
\(\boldsymbol \theta_N^{\mathcal{V}}\in\mathbb{R}^N\)), the
\(N_{\mathcal{H}}\) hidden variables \(\boldsymbol H\) (i.e.,
\(\boldsymbol \theta_N^{\mathcal{H}}\in\mathbb{R}^{N_{\mathcal{H}}}\)),
and the \(N *N_{\mathcal{H}}\) cross-product variables
\(\boldsymbol X^T\boldsymbol H\) (i.e.,
\(\boldsymbol \theta_N^{\mathcal{VH}}\in\mathbb{R}^{N*N_{\mathcal{H}}}\)).
However, unlike some previous exponential models considered in Sections
\ref{one-param-exp}-\ref{fixed-dim-exp} (cf.~Proposition
\ref{prp:prop1}, note that the RBM formulation always associates
parameters with \emph{bounded} statistics (i.e., the components of
\((\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)\)) so
that model instability cannot arise here due to the magnitude of
sufficient statistics exceeding the sample size \(N\). Instead, RBM
instability may be linked solely to parameter configuration and the fact
that the number \(q(N) \geq N\) of parameters necessarily increases with
the number \(N\) of observations \(\boldsymbol X\), in contrast to
previous exponential cases of fixed parameter dimension.
